Tidal Locking in Low-Mass Stellar Binary Stars

### 2018-10-29 Paper/sync.tex
Updates from ShareLaTeX

\shorttitle{Reproducing Kepler Stellar Rotation Periods}
\shortauthors{Fleming et al.}

@@

\newcommand{\vplanet}[0]{\texttt{VPLANET}\xspace}
\newcommand{\eqtide}[0]{\texttt{EQTIDE}\xspace}
\newcommand{\stellar}[0]{\texttt{STELLAR}\xspace}

@@

\title{Reproducing Stellar Rotation Periods in the Kepler Field via Magnetic Braking and Tidal Torques}

@@

\author{James R. A. Davenport}
\affiliation{Astronomy Department, University of Washington \\
Box 951580, Seattle, WA 98195}

\author{Rodrigo Luger}
\affiliation{Center for Computational Astrophysics, Flatiron Institute \\
New York, NY 10010}

@@

Stellar rotation period (P$_{rot}$) studies have produced numerous insights into the angular momentum evolution of low-mass stars, but previous investigations have focused on single stars even though roughly half of Sun-like stars are in stellar binaries. We examine the impact of unresolved stellar binaries on the stellar P$_{rot}$ distribution of the \textit{Kepler} field by performing Monte Carlo simulations of a population composed of single and double stars by modeling stellar evolution, magnetic braking, and tidal forces. We find that tides modify stellar rotations in all binaries with orbital periods up to 70 days.  At short orbital periods, most stellar binaries tidally-lock into synchronous rotation, naturally explaining the observed population of fast rotators in the \textit{Kepler} field that cannot be reproduced by single-star-only models.  Many binaries with longer orbital periods, some up to orbital periods of 70 days, tidally-lock, synchronizing the stellar P$_{rot}$ with the binary orbital period, causing P$_{rot}$ to not be a valid proxy for age in all cases, i.e. gyrochronology methods must be applied carefully. Our assumed flat stellar age distribution over $1-4$ Gyr underpredicts the number of fast rotators in the \textit{Kepler} field. If instead we assume a bimodal stellar age distribution in the \textit{Kepler} field composed of a young ($100$ Myr$-1$ Gyr old) and an old ($1-4$ Gyr) population, we successfully reproduce the \textit{Kepler} P$_{rot}$ distribution.

@@

Measuring stellar rotation periods via star spot modulation is a critical tool for understanding stellar angular momentum evolution in low-mass stars (M$\lsim 1.2$ M$_{\odot}$). As stars evolve, angular momentum conservation dictates that their evolving moments of inertia necessarily change their P$_{rot}$, e.g. stars spin-up as their radii contract, whereas magnetic braking, the coupling of the corotating stellar wind with the stellar magnetic field, removes angular momentum from the star, causing the observed long-term spin-down of stars \citep{Parker1958,Skumanich1972}.

Extensive measurements of ${>}10,000$ stellar rotation periods in the \textit{Kepler} field \citep{Borucki2010,McQuillan2013,Reinhold2013,McQuillan2014,Reinhold2015,Lurie2017} have provided exquisite observational constraints for theorists to model long-term stellar angular momentum evolution. Various models have been developed to explain the observed P$_{rot}$ distribution in the \textit{Kepler} field \citep[e.g.][]{Matt2015,vanSaders2018} and to link the long-term spin-down of low-mass stars to their ages, a method known as gyrochronology \citep{Skumanich1972,Barnes2003,Barnes2007,Mamajek2008,Barnes2010}. These models, however, have failed to reproduce the large observed population of rapid rotators (P$_{rot} {\lsim} 7.5$ days) across all stellar masses in the \textit{Kepler} field \citep{McQuillan2013,McQuillan2014}.  Both gyrochonology relations \citep{McQuillan2014} and a modern magnetic braking model \citep{Matt2015} suggest that these fast rotators could be at most 1 Gyr old, in tension with the expected older ages of \textit{Kepler} field stars \citep{Chaplin2014,Angus2015}.

The aforementioned models have two limitations.  First, these models assumed stellar populations composed solely of single stars, whereas roughly half of Sun-like stars are in stellar binaries \citep{Raghavan2010,Duchene2013}. Binaries are difficult to detect in photometric surveys because of low transit probabilities, so typically they are isolated using magnitude cuts \citep[e.g.][]{Davenport2018,Simonian2018}. However, 1000s of eclipsing binaries are known in the \textit{Kepler} field \citep{Kirk2016}, with likely many more unresolved binaries present in the data. In short-period binaries, tidal torques modify stellar rotations by driving the binary towards the tidally-locked state \citep{Zahn1989,Fleming2018}. Tidal forces in binaries are commonly understood to act in binaries with orbital periods (P$_{orb}$) ${\lsim} 10$ days \citep[e.g.][]{Zahn1989,Meibom2005}, however \citet{Lurie2017} measured 100s of \textit{Kepler} eclipsing binary stellar rotation periods using starspot variability and found that tides can synchronize stellar binary rotation periods up to P$_{orb} \approx 45$ days.  Recently \citet{Simonian2018} found that most rapid rotators in the \textit{Kepler} field are likely non-eclipsing, tidally-synchronized short-period binaries. Clearly, unresolved stellar binaries impart a significant signal in P$_{rot}$ distributions, particularly in the \textit{Kepler} field at short P$_{orb}$, but can also influence P$_{rot}$ measurements at larger binary P$_{orb}$.

Second, no previous study has simulated how a bimodal age distribution in the \textit{Kepler} field, suggested by \citet{McQuillan2014}, \citet{Matt2015}, and \citet{Davenport2017}, could explain the observed population of rapid rotators.  If the age distribution is in fact bimodal, a young, fast rotating population could explain the observed P$_{rot}$ distribution.  Here, we perform the first forward modeling of the angular momentum evolution of a realistic stellar population of single and double stars, accounting for stellar evolution, magnetic braking, tidal torques in stellar binaries, and testing the impact of a bimodal age distribution to reproduce the observed P$_{rot}$ distribution in the \textit{Kepler} field.

%Here, we examine the role of unresolved, tidally-interacting stellar binaries in the \textit{Kepler} field using a stellar evolution model, with magnetic braking, coupled with a secular equilibrium tidal model for stellar binaries. to see if their presence can explain features in the \textit{Kepler} P$_{rot}$ distribution. In $\S$~\ref{sec:methods} and $\S$~\ref{sec:simulations}, we describe our computational methods and simulation set-up, and discuss our results and their implications in $\S$~\ref{sec:results} and $\S$~\ref{sec:discussion}, respectively.

%% Maybe just focus on Kepler Prot observations?
% in clusters \citep[e.g. Praesepe;][]{Scholz2007,Scholz2011,Agueros2011,Delorme2011,Kovacs2014,Douglas2017} and extensive P$_{rot}$ measurements

@@

We simulate stellar rotation evolution for both single stars and stellar binaries using the coupled stellar-tidal evolution model presented in \citet{Fleming2018} and implemented in the open-source code VPLanet\footnote{VPLanet is publically available
at \href{https://github.com/VirtualPlanetaryLaboratory/vplanet}{{https://github.com/VirtualPlanetaryLaboratory/vplanet}}.} \citep[][Barnes et al., in prep]{Barnes2016}. We improve upon the stellar evolution model used in \citet{Fleming2018}, \stellar, by tracking the evolution of both the stellar radius of gyration, $r_g$, and radius, using a bicubic interpolation over mass and time of the \citet{Baraffe2015} stellar evolution models. We model stellar spin-down due to magnetic braking using the model derived in \citet{Matt2015} since it has been shown to successfully reproduce many features in the P$_{rot}$ distribution of the \textit{Kepler} field, notably the upper and lower envelopes of the latter distribution and the observed ``dip" in the distribution around M${\approx}0.6$ M$_{\odot}$. We integrate the following equation to model the net change in the stellar rotation rate due to stellar evolution and magnetic braking:

@@

For stellar binaries, we assume that they tidally interact via the ``Constant Phase Lag" (CPL) version \citep[][]{FerrazMello2008,Heller2011} of the commonly used equilibrium tidal model \citep{Darwin1880}.  The CPL model tracks the secular evolution of the binary's semi-major axis, $a$, eccentricity, $e$, and stellar rotation rates, $\omega$, due to tidal torques.  The CPL model assumes that the tidal torque on one star due to its companion arises from a linear combination of uncoupled tidal bulges, each with their own associated frequency, that maintain a fixed phase with respect to the line connecting the two stars' centers of mass and is analogous to a driven, damped harmonic oscillator \citep{Greenberg2009}. We use the \eqtide implementation of the CPL model in VPLanet described in \citet{Fleming2018}, their equations 8-16, and simultaneously integrate those equations with Eq.~(\ref{eqn:stellar_rot_rate_dt}) for stellar binaries.  All equations are integrated using a $4^{th}$ order Runge-Kutta scheme.

@@

@@

\subsection{Simulations} \label{sec:simulations}

We simulate the P$_{rot}$ distribution of the \textit{Kepler} field by performing Monte Carlo study consisting of a population of 10,000 stellar systems composed of single stars and binaries using VPLanet. For each simulation, we sample the stellar mass from a Gaussian fit to the stellar mass distribution in \citet{McQuillan2014} with $\mu=0.885$ M$_{\odot}$ and $\sigma=0.201$ M$_{\odot}$ to approximate the \textit{Kepler} mission's intended selection bias of monitoring FGK stars for Earth-like planets. We restrict stellar masses to be within [$0.1,1.2$] M$_{\odot}$ to limit the influence of subgiants since more massive stars evolve off the main sequence after $\gsim 4$ Gyr, our age limit, and because more massive stars do not experience magnetic braking due to their shallow convective envelopes \citep{Kraft1967}. Following \citet{Matt2015}, we uniformly sample the log$_{10}$ of P$_{rot}$ over [$0.8,15$] days, a distribution that approximates the P$_{rot}$ distribution of stars in the ${\sim}2$ Myr old ONC.  We randomly assign stars a stellar companion according to stellar multiplicity fractions from \citet{Duchene2013}. In binaries, we compute the secondary star's mass by uniformly sampling the mass ratio over [$0.1, 1$] following observations of mass ratios in low-mass binaries \citep{Raghavan2010}.We uniformly randomly sampling the initial eccentricity ($e$) over [$0.0,0.3$] and stellar tidal Qs are sampled from a log-uniform distribution over $[10^5,10^8]$. For the initial binary $P_{orb}$, we test two models. In Model 1, $10\%$ of binaries have their initial P$_{orb}$ uniformly sampled over [$3,80$] days, consistent with P$_{orb}$ distributions of field stars \citep[e.g.][]{Raghavan2010}, and the remaining $90\%$ have P$_{orb} > 80$ d and are considered ``single" stars since for P$_{orb} > 80$ d, tidal torques are sufficiently weak such that the stars' evolution is effectively independent of their companion.  In Model 2, we consider an enhanced population of short period binaries by sampling all initial P$_{orb}$ uniformly over [$3,80$] days.

In Models 1 and 2, we simulate each system to an age uniformly sampled over [$1,4$] Gyr, the approximate age range of \textit{Kepler} field stars \citep{McQuillan2014,Angus2015}, with the upper-limit suggested by \citet{McQuillan2014} and \citet{Matt2015}. We also consider a ``Two Population" model that uses the initial conditions of Model 2 and assumes that $40\%$ of stars comprise a ``young" population with ages uniformly sampled over [$0.1,1$] Gyr, while the remainder comprise the ``old" population with ages uniformly sampled over [$1,4$] Gyr. We track the P$_{rot}$ evolution for each single star and the primary, i.e. more massive star, in binaries as it is observationally easier to measure a P$_{rot}$ due to starspot modulations on the more massive, and hence brighter, star. We compare our simulation results with the \textit{Kepler} P$_{rot}$ distribution measured by \citet{McQuillan2014}.

All code used to run simulations and generate figures is available online\footnote{\href{https://github.com/dflemin3/binGyro}{https://github.com/dflemin3/binGyro}.} with accompanying descriptions and documentation.

%  For simulations of binaries, given the inherent uncertainty in and complexity of the formation of binaries \citep[e.g.][]{Bonnell1994,Bate2000,Bate2002,Moe2018} and the potential for dynamical processing via tides or stellar close encounters \citep[e.g.][]{Mardling2001,Hurley2002,Ivanova2005,Meibom2005}, we take an agnostic approach to the initial orbital configuration.

%Our goal is to determine how the distribution of $P_{rot}$ for a population stars composed solely of single stars differs, if at all, from a population of both single stars and binaries.  If there is a substantial difference between the two distributions, e.g. tidally-locked fast rotators, then any study examining the distribution of stellar $P_{rot}$ should therefore account for the impact of tidally-interacting stellar binaries.  We approach this problem with a ...

@@

\subsection{Tidal Locking in Stellar Binaries} \label{sec:results:locking}

Here we examine the tidal locking in binaries using the simulations from Model 2 (see $\S$~\ref{sec:results:kepler2}) by plotting P$_{rot}$ as a function of mass and age for binaries in Fig.~\ref{fig:locking}. We find that $35.5\%$ of binaries tidally lock before the age of the system, typically for tidal $Q \lsim 10^7$. Tidally-locked binaries tend to rotate more rapidly than unlocked binaries, reflected in the increased density for P$_{rot} \lsim 20$ d in the marginal distribution for the locked population and in the distribution medians (locked median P$_{rot} = 19.8$ d compared to unlocked median P$_{rot} = 27.5$ d). Binaries readily tidally-lock at short P$_{orb}$ due to the tidal force's strong semi-major axis dependence, increasing the strength of tidal torques at shorter orbital separations. When stellar binaries approach the tidally-locked state, tidal torques overpower the torque due to magnetic braking, fixing P$_{rot} = $P$_{orb}$, resulting in long-lasting rapid rotation in short P$_{orb}$ binaries.

Tidal forces modify the spins of unlocked binaries as well, tending to spin-up stars against the effects of magnetic braking towards orbital synchronization, reflected in unlocked binaries' median P$_{rot} = 27.5$ d, compared to the single star median P$_{rot} = 30.6$ d. Tidal locking is not limited to P$_{orb} \lsim 10$ d, however, as we observe stellar binaries to tidally lock out up to P$_{orb} \approx 70$ d, producing a slow-rotating population above the P$_{rot}$ distribution envelop of single stars (compare to Fig.~\ref{fig:kepler2}). This behavior is consistent with observations of P$_{rot}$ in \textit{Kepler} eclipsing binaries by \citet{Lurie2017} who find that binaries can tidally lock up to their detection limit of P$_{orb} = $ P$_{rot} = 45$ d. The observed age distribution of tidally-locked binaries has little dependence on P$_{rot}$, with both young and old systems tidally-locking at long and short P$_{rot}$, causing the predictions of gyrochronology models to fail in such systems.

\begin{figure}[h]
	\includegraphics[width=0.45\textwidth]{../Plots/locked_kepler.pdf}
   \caption{Simulated distribution of P$_{rot}$ for tidally-locked (cross) and unlocked (circle) stellar binaries, colored by age for Model 2 (see $\S$~\ref{sec:results:kepler2}). Left: P$_{rot}$ as a function of stellar mass.  Right: Marginalized P$_{rot}$ distribution for locked (blue) and unlocked (orange) binaries.  Tidally-locked binaries rotate more rapidly than unlocked binaries, displaying an increased density in the P$_{rot}$ distribution for P$_{rot} {\lsim} 20$ d, and with a median P$_{rot} = 19.8$ d, compared to the unlocked median P$_{rot} = 27.5$ d. Unlocked binaries on average rotate more quickly than single stars (compare to single distribution in Fig.~\ref{fig:kepler2}) due to the influence of tidal torques.}
    \label{fig:locking}
\end{figure}

\subsection{Comparison to \textit{Kepler}}

\subsubsection{Model 1} \label{sec:results:kepler1}

In Fig.~\ref{fig:kepler1}, we display our simulated P$_{rot}$ for Model 1, compared to the observed \textit{Kepler} distribution \citep{McQuillan2014}. Single stars, defined as stars without a companion, or with a companion with a P$_{orb} > 80$ d where tides do not modify the orbit or stellar rotations, rotate more slowly with age due to magnetic braking, with more massive stars rotating more quickly than low mass stars, consistent with \textit{Kepler} P$_{rot}$ observations.  There is a larger spread in P$_{rot}$ for stars with M$\lsim 1$ M$_{\odot}$ with the spread increasing with decreasing stellar mass, due to the mass-dependence magnetic braking torque in the \citet{Matt2015} model.
%in which lower mass stars tend to spin down more quickly than higher-mass stars once they enter the "unsaturated" regime (see \citet{Matt2015} equations 6 and 7).

The inclusion stellar binaries naturally produces a population of fast rotators (P$_{rot} < 7.5$ d) across all masses as binaries readily tidally-lock at shorter P$_{orb}$, with ${\sim}90\%$ of stars in this P$_{rot}$ regime being tidally-locked binaries. Single-star only models fail to reproduce this feature.  The fast rotator population only comprises $0.25\%$ of all simulated stars, compared to $16\%$ of \textit{Kepler} stars in the same regime, and is therefore insufficient to explain the observed population of fast rotators. Our simulated P$_{rot}$ distribution is systematically biased towards longer P$_{rot}$ than the \textit{Kepler} distribution with a total median P$_{rot} = 30.3$ d compared to \textit{Kepler} median P$_{rot} = 17.3$ d. The inclusion of binaries does little to shift the P$_{rot}$ distribution towards shorter P$_{rot}$ as Model 1's assumed binary P$_{orb}$ distribution, based on observations of field stars \citep{Raghavan2010}, predicts that only ${\sim}4\%$ of all stars will be in stellar binaries with P$_{orb} < 80$ d, short enough for tides to appreciably modify stellar rotations.

\begin{figure*}[ht]
	\includegraphics[width=\textwidth]{../Plots/kepler_box_dist_corr.pdf}
   \caption{Simulated distribution of P$_{rot}$ for single (circle) and double (cross, defined as stellar binaries with P$_{orb} < 80$ d) stars for Model 1. Left: P$_{rot}$ as a function of stellar mass, colored by age.  Center: Marginalized P$_{rot}$ distributions for single stars (orange), binary stars (green), and the total sample (blue), and \textit{Kepler} \citep{McQuillan2014} (red). Except for a few low-mass (${\lsim}0.3$M$_{\odot}$) single stars, regardless of age, rotate more slowly (single median P$_{rot} = 30.5$ d) than \textit{Kepler} stars (\textit{Kepler} median P$_{rot} = 17.3$ d) and fail to explain the observed population of fast rotators.  Binaries, however, readily tidally-lock at shorter P$_{orb}$, naturally producing a population of fast rotators (binary median P$_{rot} = 25.9$ d). Right: box-and-whisker plot of the P$_{rot}$ distribution for the simulations seen in the left panel in 0.1 M$_{\odot}$ bins. The shaded regions cover the distributions' interquartile range, the white lines indicate the medians, and the whiskers do not indicate error bars, but rather the spread of the distribution, extending to the 5th and 95th percentiles. The numbers over the boxes indicate the percent difference between the single (orange) and total (blue) distribution medians with the \citet{McQuillan2014} (red) sample. Even with the inclusion of binaries that tidally-lock at short P$_{rot}$, Model 1 produces a more slowly rotating stellar population than is observed in the \textit{Kepler} field, indicating that our assumed stellar population is incorrect.}%
    \label{fig:kepler1}%
\end{figure*}

\subsubsection{Model 2: Enhanced Binary Population} \label{sec:results:kepler2}

In Model 2, we attempt to increase the number of fast rotators by increasing the number of systems that could tidally-lock at short P$_{rot}$.  We assume all stellar binaries have a P$_{orb}$ uniformly sampled over [$3,80$] d, as opposed to just $10\%$ of binaries in Model 1.  Although optimistic, our assumption is reasonable as an analysis of fast rotators in the \textit{Kepler} field by \citet{Simonian2018} found that $\sim 59\%$ of stars with $1.5 <$ P$_{rot} < 7$ d are likely unresolved, tidally-locked binaries, suggesting that a substantial short-period rapidly rotating binary population does exist in the \textit{Kepler} field.

As in Model 1, stellar binaries tidally-lock at short P$_{rot}$ producing a population of fast rotators below the single star sequence with binaries comprising ${\sim}99\%$ of fast rotators.  The inclusion of additional short P$_{orb}$ binaries noticeably modifies the simulated P$_{rot}$ distribution, shifting the median P$_{rot}$ for the total sample down to $28.6$ d, compared to $30.6$ d for the single-star-only distribution.  The impact of binaries is evident across all masses in the right panel of Fig.~\ref{fig:kepler2} as the total P$_{rot}$ distribution, which includes binaries, better matches the observed \textit{Kepler} distribution than the single star only distribution, due to the fast rotating tidally-locked binary population.  The enhanced binary population of Model 2 still predicts a more slowly rotating stellar P$_{rot}$ distribution that is observed, specifically in the fast rotator regime, with Model 2 predicting $3.4\%$ of our simulated stars are fast rotators, compared to $16\%$ of \textit{Kepler} stars. As noted in $\S$~\ref{sec:results:locking}, tidally-locked binaries fix P$_{rot} = $ P$_{orb}$ as tidal torques overpower the torque due to magnetic braking, causing P$_{rot}$ in binaries to not correlate with age.

>>>>

\shorttitle{Tidal Locking in Stellar Binaries}
\shortauthors{Fleming \& Barnes}

@@

\newcommand{\vplanet}[0]{\texttt{VPLanet}\xspace}
\newcommand{\eqtide}[0]{\texttt{EQTIDE}\xspace}
\newcommand{\stellar}[0]{\texttt{STELLAR}\xspace}
\newcommand{\kepler}[0]{\textit{Kepler}\xspace}

@@

\title{Tidal Locking in Low-Mass Stellar Binary Stars}

@@

@@

Abstract

@@

XXX

tidal locking assumed for short period binaries, tsync << tcirc eg Mazeh 2008, so some have speculated that low-mass binaries can synchronize out to longer orbital periods.  tentative evidence for this in Lurie 2017 data, discuss pseudosyncronization, synchronization, \citet{Counselman1973} tidal end states

we give a shit about sync binaries as they are fast rotators, e.g. what \citet{Simonian2018} might be seeing in \kepler field, can have enhanced magnetic activity (short Prot -> enhanced Lxuv), impact gyrochronology predictions, it's a fundamental evolutionary process for roughly half of sun like stars.

dank plot showing average tsync for various orbital period ranges for CPL vs CTL marginalizing over all other parameters. do it with small multiples

@@

XXX

We simulate stellar rotation evolution for both single stars and stellar binaries using the coupled stellar-tidal evolution model presented in \citet{Fleming2018} and implemented in the open-source code VPLanet\footnote{VPLanet is publically available
at \href{https://github.com/VirtualPlanetaryLaboratory/vplanet}{{https://github.com/VirtualPlanetaryLaboratory/vplanet}}.} \citep[][Barnes et al., in prep]{Barnes2016,vplanet2018}. We improve upon the stellar evolution model used in \citet{Fleming2018}, \stellar, by tracking the evolution of both the stellar radius of gyration, $r_g$, and radius, using a bicubic interpolation over mass and time of the \citet{Baraffe2015} stellar evolution models. We model stellar spin-down due to magnetic braking using the model derived in \citet{Matt2015} since it has been shown to successfully reproduce many features in the P$_{rot}$ distribution of the \textit{Kepler} field, notably the upper and lower envelopes of the latter distribution and the observed ``dip" in the distribution around M${\approx}0.6$ M$_{\odot}$. We integrate the following equation to model the net change in the stellar rotation rate due to stellar evolution and magnetic braking:

@@

 Equilibrium tidal models, first introduced by \citep{Darwin1880}, track the secular evolution of an orbiter's semi-major axis, $a$, eccentricity, $e$, and the rotation rates, $\omega$, and obliquities $\psi$, of the tidally-interacting bodies subject to tidal torques. Equilibrium tidal models assume that tidally-interacting bodies raise tidal bulges on their companions that are offset from the line connecting the both bodies' centers of mass, driving torques that exchange angular momentum between the orbit and both bodies' spins. Equilibrium tidal models are linear since the tidal waves that comprise the tidal bulge on a body are assumed to be uncoupled. Under these assumptions, the tidal evolution is analogous to a driven, damped harmonic oscillator \citep{Greenberg2009}.  We refer the reader to \citet{Barnes2017} for an in-depth discussion of the assumptions and limitations of equilibrium tidal models.  Although simple, equilibrium tidal models have been used to robustly model the secular orbital and rotation evolution of both exoplanets \citet[e.g.][]{Goldreich1966,Jackson2009,Leconte2010,Heller2011,Barnes2013,Barnes2017} and stellar binaries \citep[e.g.][]{Repetto2014,Fleming2018}.  Here, we consider two equilibrium tidal models to study the secular spin-orbital evolution of low-mass stellar binaries.

\subsubsection{Constant Phase Lag (CPL) Model}

The ``Constant Phase Lag" (CPL) \citep[][]{FerrazMello2008,Heller2011} equilibrium tidal model assumes that the tidal torque on one body due to its companion arises from a linear combination of several discrete, uncoupled tidal bulges, each with their own associated frequency, that maintain a fixed phase with respect to the line connecting the two stars' centers of mass. We use the \eqtide implementation of the CPL model in VPLanet following the derivation of \citet{FerrazMello2008}.  The equations that govern the secular change in $e$ and $a$ are as follows:

\begin{equation} \label{eqn:cpl_e}
\frac{de}{dt} = -\frac{ae}{8 G m_1 m_2} \sum_{i=1}^2 Z_{i,\mathrm{CPL}} \left( 2 \varepsilon_{0,i} - \frac{49}{2} \varepsilon_{1,i} + \frac{1}{2} \varepsilon_{2,i} + 3 \varepsilon_{5,i} \right)
\end{equation}
\begin{equation} \label{eqn:cpl_a}
\frac{da}{dt} = \sum_{i=1}^2 \frac{da_i}{dt}
\end{equation}
where if the $i^{th}$ body is tidally-locked in a synchronous orbit,
\begin{equation} \label{eqn:cpl_dadt_locked}
\frac{da_{i,sync}}{dt} = -\frac{a^2}{G m_1 m_2} Z_{i,\mathrm{CPL}} \left( 7 e^2 + \sin^2 (\psi_i) \right) \varepsilon_{2,i},
\end{equation}
otherwise
\begin{equation}
\begin{split}
\frac{da_i}{dt} & = \frac{a^2}{4 G m_1 m_2} Z_{i,\mathrm{CPL}} \left( 4 \varepsilon_{0,i} + e^2 \left[ -20 \varepsilon_{0,i} + \frac{147}{2} \varepsilon_{1,i} \right. \right. \\
&  + \left. \left. \frac{1}{2} \varepsilon_{2,i} - 3 \varepsilon_{5,i} \right] - 4 \sin^2 (\psi_i) \left[ \varepsilon_{0,i} - \varepsilon_{8,i} \right] \right).
\end{split}
\end{equation}
The CPL equations for $\psi$ and $\omega$ evolution are
\begin{equation} \label{eqn:cpl_psi}
\frac{d\psi_i}{dt} = \frac{Z_{i,\mathrm{CPL}} \sin(\psi_i)}{4 m_i r_{g,i}^2 R_i^2 n \omega_i} \left( [1-\xi_i] \varepsilon_{0,i} + [1+\xi_i](\varepsilon_{8,i} - \varepsilon_{9,i}) \right)
\end{equation}
\begin{equation} \label{eqn:cpl_omega}
\begin{split}
\frac{d\omega_i}{dt}& = -\frac{Z_{i,\mathrm{CPL}}}{8m_i r_{g,i}^2 R_i^2 n} \left(4 \varepsilon_{0,i} + e^2\left[-20\varepsilon_{0,i} + 49\varepsilon_{1,i} + \varepsilon_{2,i} \right] \right. \\
& \left. + 2 \sin^2(\psi_i) \left[ -2 \varepsilon_{0,i} + \varepsilon_{8,i} + \varepsilon_{9,i} \right] \right)
\end{split}
\end{equation}
where $G$ is Newton's gravitational constant, $n$ is the binary's mean motion, and the index $i$ denotes that $i^{th}$ body. The tidal phase lags signs, $\varepsilon$, for the $i^{th}$ body are given by
\begin{equation} \label{eqn:cpl_eps}
\begin{split}
\varepsilon_{0,i} & = \Sigma(2 \omega_i - 2n) \\
\varepsilon_{1,i} & = \Sigma(2 \omega_i - 3n) \\
\varepsilon_{2,i} & = \Sigma(2 \omega_i - n) \\
\varepsilon_{5,i} & = \Sigma(n) \\
\varepsilon_{8,i} & = \Sigma(\omega_i - 2n) \\
\varepsilon_{9,i} & = \Sigma(\omega_i)
\end{split}
\end{equation}
where the function $\Sigma(x)$ returns $1$ for positive $x$, $-1$ for negative $x$, and $0$ otherwise.

The intermediate variable $Z_{\mathrm{CPL},i}$ is given by
\begin{equation} \label{eqn:cpl_z}
Z_{i,\mathrm{CPL}} = 3 G^2 k_{2,i} m_j^2 (m_1 + m_2) \frac{R_i^5}{a^9} \frac{1}{n Q_i}
\end{equation}
where the $j^{th}$ body is the $i^{th}$ body's companion, $k_{2}$ is the body's Love number of degree 2, and $Q$ is the tidal quality factor (``tidal Q"). The tidal Q parameterizes the energy dissipation due to tidal evolution, with lower tidal Qs, i.e. larger phase differences between the tidal bulges, driving more rapid tidal evolution.

$\xi_i$ is defined as
\begin{equation}\label{eqn:cpl_chi}
\xi_i = \frac{r_{\mathrm{g},i}^2 R_i^2 \omega_i a n }{ G M_j}.
\end{equation}

@@

\subsubsection{Constant Time Lag (CTL) Model}

The ``Constant Time Lag" (CTL) \citep[][]{Hut1981,Leconte2010} equilibrium tidal model assumes a constant time interval between the body's tidal bulge and the passage of the tidally-interacting companion. In this formalism, unlike the CPL model, the CTL model is continuous over a range of tidal wave frequencies.  However if the assumption of linearity is relaxed, i.e. frequencies associate with tidal bulges are allowed to depend on a spin or orbital forcing frequency, then this model is only valid over a small range of frequencies \citep{Greenberg2009}. We use the \eqtide implementation of the CTL model in VPLanet following the derivation of \citet{Leconte2010}.  The equations that govern the secular change in $e$, $a$, $\omega$, and $\psi$ are as follows:

\begin{equation} \label{eqn:ctl_e}
  \frac{de}{dt} = \frac{11 ae}{2 G M_1 M_2}
  \sum_{i = 1}^2 Z_{\mathrm{CTL},i} \left( \cos(\psi_i) \frac{f_4(e)}{\beta^{10}(e)}  \frac{\omega_i}{n} -\frac{18}{11} \frac{f_3(e)}{\beta^{13}(e)}\right),
\end{equation}

\begin{equation}\label{eqn:ctl_a}
  \frac{da}{dt} \ = \  \frac{2 a^2}{G M_1 M_2}
  \sum\limits_{i = 1}^2 Z_{\mathrm{CTL},i} \left( \cos(\psi_i) \frac{f_2(e)}{\beta^{12}(e)} \frac{\omega_i}{n} - \frac{f_1(e)}{\beta^{15}(e)}\right),
\end{equation}

\begin{equation}\label{eqn:ctl_o}
  \frac{d\omega_i}{dt} \ = \ \frac{Z_{\mathrm{CTL},i}}{2 M_i r_{g,i}^2 
R_i^2 n} \left( 2 \cos(\psi_i) \frac{f_2(e)}{\beta^{12}(e)} - \left[ 1+\cos^2(\psi)
 \right] \frac{f_5(e)}{\beta^9(e)} 
\frac{\omega_i}{n} \right),  
\end{equation}
and
\begin{equation}\label{eqn:ctl_psi}
  \frac{d\psi_i}{dt} = \frac{Z_{\mathrm{CTL},i} \sin(\psi_i)}{2 M_i r_{g,i}^2 R_i^2 n \omega_i}\left( \left[ \cos(\psi_i) - \frac{\xi_i}{ \beta} \right] \frac{f_5(e)}{\beta^9(e)} \frac{\omega_i}{n} - 2 \frac{f_2(e)}{\beta^{12}(e)} \right).
\end{equation}
where the intermediate variables are given by 
\begin{equation}\label{eqn:ctl_z}
 Z_{i,\mathrm{CTL}} = 3 G^2 k_{2,i} M_j^2 (M_i+M_j) \frac{R_i^5}{a^9} \tau_i ,
\end{equation}
and 
\begin{equation}\label{eqn:ctl_f_e}
\begin{array}{l}
\beta(e) = \sqrt{1-e^2},\\
f_1(e) = 1 + \frac{31}{2} e^2 + \frac{255}{8} e^4 + \frac{185}{16} e^6 + \frac{25}{
64} e^8,\\
f_2(e) = 1 + \frac{15}{2} e^2 + \frac{45}{8} e^4 + \frac{5}{16} e^6,\\
f_3(e) = 1 + \frac{15}{4} e^2 + \frac{15}{8} e^4 + \frac{5}{64} e^6,\\
f_4(e) = 1 + \frac{3}{2} e^2 + \frac{1}{8} e^4,\\
f_5(e) = 1 + 3 e^2 + \frac{3}{8} e^4.
\end{array}
\end{equation}

In both the CPL and CTL model, We assume $k_2 = 0.5$. This choice of $k_2$ does not impact our results as $k_2$ is degenerate with Q in the CPL model, e.g. the $k_2/Q$ scaling in Eq.~(\ref{eqn:cpl_z}), and with $\tau$ in the CTL model, e.g. $k_2 \tau$ scaling in Eq.~(\ref{eqn:ctl_z}), so we instead examine how our results scale with $Q$ and $\tau$.

\subsubsection{Tidal Locking and Numerical Integration}

XXX discuss integration specifics, equilbrium rotation rates, timestepping, $4^{th}$ order Runge-Kutta scheme.

\subsubsection{Example Tidal Evolution}

Here in Fig.~\ref{fig:tidalExample} we plot the example tidal evolution for $a$, $e$, and $\omega$, ignoring stellar evolution, for a solar-twin binary with an initial P$_{orb} = 10$ d, P$_{rot} = 1$ d, $e = 0.2$ for the CPL model and CTL model, assuming $Q=10^6$ and $\tau = 0.1$ seconds, respectively.

\begin{figure*}[h]
	\includegraphics[width=\textwidth]{../Plots/tidalExample.pdf}
   \caption{Spin-orbital tidal evolution of a 1 M$_{\odot} -$ 1 M$_{\odot}$ stellar binary. XXX}%
    \label{fig:tidalExample}%
\end{figure*}

@@

\section{Simulations} \label{sec:simulations}

XXX

- we consider eccentricities up to 0.3 as the CPL model fails above that.  CTL model can handle it, but at high eccentricity, dynamical tide, e.g. dissipation via waves and stuff, becomes important

@@

XXX

%\begin{figure*}[ht]
%	\includegraphics[width=\textwidth]{../Plots/kepler_two_pop_box_dist.pdf}
%   \caption{Identical format to Fig.~\ref{fig:kepler1}, but displaying simulated data from our two population model (see $\S$~\ref{sec:results:kepler_two_pop}). Our two population model reproduces the observed \textit{Kepler} P$_{rot}$ distribution, especially the fast rotator population, with a total median P$_{rot} = 19.6$ d compared to the \textit{Kepler} median P$_{rot} = 17.3$ d, a close match. In the right panel for $ 0.4 \textrm{M}_{\odot} \lsim \textrm{M} \lsim 1.2 \textrm{M}_{\odot}$, the inclusion of tidally-interacting binaries in the total (blue) sample leads to an excellent match with the \textit{Kepler} distribution (red) compared to the single-star only (orange) model, emphasizing the importance of including stellar binaries in P$_{rot}$ forward models.}%
%    \label{fig:kepler_two_pop}%
%\end{figure*}
